Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon

### 2022-06-12 analysis/paper/paper.Rmd
minor edits from TAFS review

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is within a section of the Salmon River containing historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon populations, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015], and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

@@

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (rkm) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. A large alluvial fan at the mouth of Dump Creek has, at least partially, created a hydraulic control in the Salmon River resulting in the formation of Deadwater Slough. Although the origin and timing of the alluvial fan and Deadwater Slough is somewhat ambiguous [@Reichmuth1985; @USACE1986], some believe they coincided with the failure of a small mining reservoir in the Dump Creek drainage around 1897 that deposited substantial amounts of sediment at its confluence with the Salmon River [@Emerson1973]. The slough is currently a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and has characteristics resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Our intent was to survey during the emigration of the 2018 Chinook Salmon brood year that occurred fall 2019 and spring 2020, however logistical constraints during the onset of the COVID-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. Fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration and emigration.

@@

Using hook-and-line angling, we caught a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` Northern Pikeminnow over the course of the study including `r sum(N_mods[[3]][[1]]$m) + sum(N_mods[[3]][[3]]$m)` recaptures; the overall CPUE was `r round(sum(cpue_df$n_fish) / (sum(cpue_df$total_effort_min) / 60), 2)` Northern Pikeminnow per angler hour (Table \@ref(tab:summary-tbl)). Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple census estimator compared to the single census and adjusted delayed-mixing multiple census estimators (Figure \@ref(fig:abund-fig)). Our sampling design most closely matched a multiple census estimator; therefore, the Schnabel estimates were considered most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r prettyNum(round(mean(all_est$Lci[all_est$model == "Schnabel"])), big.mark = ",")` - `r prettyNum(round(mean(all_est$Uci[all_est$model == "Schnabel"])), big.mark = ",")`) using the unadjusted Schabel estimator. Using the delayed-mixing Schnabel estimator the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). All subsequent analyses use results from the adjusted delayed-mixing Schnabel estimator. For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish per 100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

During the fall DSR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consisted of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet was fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Daily consumption rates were higher early in the fall and late in the spring and corresponded with higher water temperatures, especially above 8 degrees Celsius (Figure \@ref(fig:bio-p)). Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` in the fall and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population was estimated to consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

Although the origins of Deadwater Slough are somewhat ambiguous [@Reichmuth1985; @USACE1986], it appears likely that human activities in Dump Creek have either exacerbated or contributed to its formation, and in turn, created favorable conditions for Northern Pikeminnow. Among the three surveys, our lowest population abundance estimate was greater than 10,000 Northern Pikeminnow occupying Deadwater Slough, suggesting a remarkably high density given the size of the area. The slow water velocity and lack of cover for fish also create conditions where juvenile salmonids that are rearing in or emigrating through Deadwater Slough, including Chinook Salmon, are susceptible to predation. We estimate that, at a minimum, Northern Pikeminnow in Deadwater Slough consume greater than `r prettyNum(round(min(total_impacts$juv_Lci), -3), big.mark = ",")` juvenile Chinook Salmon (the lower endpoint of the confidence interval for the most conservative scenario); however, it is likely that this value is closer to `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50], -3), big.mark = ",")`, annually. The magnitude of predation by Northern Pikeminnow is equivalent to `r round((min(total_impacts$adult_N) / 558) * 100)`% and `r round((max(total_impacts$adult_N) / 558) * 100)`% of recent adult escapements. Deadwater Slough predates the ESA-listing of Chinook Salmon populations in the Upper Salmon MPG in the 1990s and is therefore unlikely to be the primary cause for the population's decline. However, predation by Northern Pikeminnow in Deadwater Slough and elsewhere should be considered among limiting factors hindering recovery efforts of ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

@@

```{r bio-tbl}
bio_tbl = bio_g_summ %>%
  mutate(`Date Range` = season) %>%
  select(season,
         `Date Range`,
         name,
         cumm_g) %>%
  mutate(name = str_sub(name, 2, 3)) %>%
  rename(Season = season,
         `Fish in Diet (%)` = name,
         `Fish Consumed (g)` = cumm_g) %>%
  mutate(`Date Range` = recode(`Date Range`,
                               Fall = "Sep 15 - Nov 30",
                               Spring = "Mar 1 - May 31")) %>%
  kable(booktabs = T,
        caption = "Estimated fish consumed by an individual Northern Pikeminnow to maintain its body weight during the fall DSR and spring NRR Chinook Salmon emigrations. Bioenergetic model results for each season include seven model runs that assume different percentages of fish in the diet. Northern Pikeminnow start and end weights were 598.7 and 430.7 g for the fall and spring seasons, respectively.") %>%
  kable_styling()

>>>>

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of uncharacteristically slow and deep water in the Salmon River, Idaho. The Deadwater Slough is within a section of the Salmon River containing historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon populations, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015], and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

@@

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (rkm) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. A large alluvial fan at the mouth of Dump Creek has, at least partially, created a hydraulic control in the Salmon River resulting in the formation of Deadwater Slough. Although the origin and timing of the alluvial fan and Deadwater Slough is somewhat ambiguous [@Reichmuth1985; @USACE1986], some believe they coincided with the failure of a small mining reservoir in the Dump Creek drainage around 1897 that deposited substantial amounts of sediment at its confluence with the Salmon River [@Emerson1973]. The slough is currently a slow, deep (approximately 6 m maximum depth) section in the river, spanning `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and has characteristics resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Our intent was to survey during the emigration of the 2018 Chinook Salmon brood year that occurred fall 2019 and spring 2020, however logistical constraints during the onset of the COVID-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. Fall surveys were constrained to two weeks to minimize the effect of Northern Pikeminnow immigration and emigration on the population size estimate.

@@

Using hook-and-line angling, we caught a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` Northern Pikeminnow over the course of the study including `r sum(N_mods[[3]][[1]]$m) + sum(N_mods[[3]][[3]]$m)` recaptures; the overall CPUE was `r round(sum(cpue_df$n_fish) / (sum(cpue_df$total_effort_min) / 60), 2)` Northern Pikeminnow per angler hour (Table \@ref(tab:summary-tbl)). Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple census estimator compared to the single census and adjusted delayed-mixing multiple census estimators (Figure \@ref(fig:abund-fig)). Our sampling design most closely matched a multiple census estimator; therefore, the Schnabel estimates were considered most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events (2019 and 2020) was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r prettyNum(round(mean(all_est$Lci[all_est$model == "Schnabel"])), big.mark = ",")` - `r prettyNum(round(mean(all_est$Uci[all_est$model == "Schnabel"])), big.mark = ",")`) using the unadjusted Schabel estimator. Using the delayed-mixing Schnabel estimator the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). All subsequent analyses use results from the adjusted delayed-mixing Schnabel estimator. For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish per 100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

During the fall DSR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish across the season to maintain their body size, assuming 60% of their diet consisted of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet was fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Daily consumption rates were higher early in the fall and late in the spring and corresponded with higher water temperatures, especially above 8 degrees Celsius (Figure \@ref(fig:bio-p)). Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` in the fall and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population was estimated to consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

Although the origins of Deadwater Slough are somewhat ambiguous [@Reichmuth1985; @USACE1986], it appears likely that human activities in Dump Creek have either exacerbated or contributed to its formation, and in turn, created favorable conditions for Northern Pikeminnow. Among the three surveys, our lowest population abundance estimate was greater than 10,000 Northern Pikeminnow occupying Deadwater Slough, suggesting a remarkably high density given the size of the area. The slow water velocity and lack of cover for fish also create conditions where juvenile salmonids that are rearing in or emigrating through Deadwater Slough, including Chinook Salmon, are susceptible to predation. We estimate that, at a minimum, Northern Pikeminnow in Deadwater Slough consume greater than `r prettyNum(round(min(total_impacts$juv_Lci), -3), big.mark = ",")` juvenile Chinook Salmon (the lower endpoint of the confidence interval for the most conservative scenario); however, it is likely that this value is closer to `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50], -3), big.mark = ",")`, annually. The magnitude of predation by Northern Pikeminnow on adult returns is equivalent to `r round((min(total_impacts$adult_N) / 558) * 100)`% and `r round((max(total_impacts$adult_N) / 558) * 100)`% of recent adult escapements. Deadwater Slough predates the ESA-listing of Chinook Salmon populations in the Upper Salmon MPG in the 1990s and is therefore unlikely to be the primary cause for the population's decline. However, predation by Northern Pikeminnow in Deadwater Slough and elsewhere should be considered among limiting factors hindering recovery efforts of ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

@@

```{r bio-tbl}
bio_tbl = bio_g_summ %>%
  mutate(`Date Range` = season) %>%
  select(season,
         `Date Range`,
         name,
         cumm_g) %>%
  mutate(name = str_sub(name, 2, 3)) %>%
  rename(Season = season,
         `Fish in Diet (%)` = name,
         `Fish Consumed (g)` = cumm_g) %>%
  mutate(`Date Range` = recode(`Date Range`,
                               Fall = "Sep 15 - Nov 30",
                               Spring = "Mar 1 - May 31")) %>%
  kable(booktabs = T,
        caption = "Estimated fish consumed by an individual Northern Pikeminnow to maintain its body weight during the fall DSR and spring NRR Chinook Salmon emigrations. Bioenergetic model results for each season include seven model runs that assume different percentages of fish in the diet. Northern Pikeminnow start and end weights were 598.7 g for the fall season and 430.7 g for the spring season.") %>%
  kable_styling()
